---
title: "Computing Intron Retention With KeepMeAround (`kma`)"
author: "Harold Pimentel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette described how to run the Keep Me Around `kma` suite to compute
intron retention in RNA-Seq experiments.

# General pipeline

The general pipeline is as follows:

1. **Pre-process** to generate intron coordinates and sequences
1. **Quantify** transcript expression using
   [eXpress](http://bio.math.berkeley.edu/eXpress/) against augmented
   transcriptome containing introns
1. **Post-process** to compute intron retention using `R` package

**TODO: Put a more detailed flowchart here**

## Installing

There are two required portions of the tool to perform intron retention
quantification, though installing the post-processing tools (intron retention)
installs the python code with the exception of dependencies. The
**pre-process** step is written in python and only needs to be run when the
annotation changes. The **post-processing** step is written in `R` and where
most of your interaction will take place.

### Installing the pre-processing tools

If you have installed the `R` package (see [Installing the post-processing
tools](#installing-the-post-processing-tools)), then you have successfully
installed the pre-processing tools. From `R`, you can find the path by typing:

```{r}
system.file("pre-process", package="kma")
```

The only additional dependencies are the Python packages `pyfasta`, `biopython` and `pysam`.
All packages can be installed via PyPI:

```{bash}
pip install pyfasta biopython pysam
```

### Installing the quantification tools

The tools needed for quantification are:

- [Bowtie 2](http://bowtie-bio.sourceforge.net/bowtie2/) short read aligner
- [eXpress](http://bio.math.berkeley.edu/eXpress/) RNA-Seq quantification tool

Please see the corresponding documentation on their respective web pages.

### Installing the post-processing tools

The pre- and post-processing tools are contained in an `R` packaged called
`kma`.  The development version is always on
[Github](http://github.com/pachterlab/kma). To install, `kma` and required `R`
dependencies, runs the code below:

```{r, eval=FALSE}
required_packages <- c("devtools", "data.table", "reshape2", "dplyr")
install.packages(required_packages)
devtools::install_github("http://github.com/pachterlab/kma")
```

You can then load the `kma` R package like you would any other R packages:

```{r}
library("kma")
```

## Organization
We recommend organizing your experiments in the following format:

```
experiment
|____conditionA
| |____conditionA1
| |____conditionA2
| |____conditionA3
|____conditionB
| |____conditionB1
| |____conditionB2
| |____conditionB3
```

where `experiment` is the top level for the particular set of experiments,
`conditionX` refers to the condition (e.g. tumor, or control), and
`conditionXY` represents the biological replicates. This allows for
a structured way to keep track of your data as well as an easy way to load it
in `R`. An example can be seen in the [worked example](#a-worked-example)
section.

## Generate intron coordinates (pre-processing)

**Note**: Pre-processing only has to be run when the gene annotation changes.
If you use the same annotation for many experiments, you won't be running this
tool often.

The path to the pre-processing tools can be found in R by typing
`system.file("preprocess", package="kma")`. Open a terminal and put this path
in an environment variable:

```{bash}
PRE=/path/to/kma/preprocess
```

If you are on a Mac, the path might look like: /Library/Frameworks/R.framework/Versions/3.1/Resources/library/kma/preprocess

We are now ready to generate the introns. A typical command line call to the
pre-processing tools looks like:

```{bash}
python $PRE/generate_introns.py --genome seq.fa --gtf trans.gtf --extend N --out out_dir
```

Where the inputs are:

- `--genome seq.fa`: **genome sequence**  in [Multi-FASTA
  format](http://en.wikipedia.org/wiki/FASTA_format) where contig names
  correspond to the GTF file.
- `--gtf trans.gtf`: **annotation file** in [GTF
  format](http://www.ensembl.org/info/website/upload/gff.html).
- `--extend N`: Optional, though we strongly recommend setting it. If set, `N`
  is the number of bases bases to overlap into the intron. Note, this should be
  at most `read_length - 1`, but we suggest making it smaller.
- `--out out_dir`: A directory to write the outputs. Will be created if doesn't
  already exist.

The following outputs will then be put in `out_dir`:

- **introns.fa** - a FASTA file containing the intron sequences.
- **introns.bed** - BED file with coordinates used to quantify intron.
- **intron\_to\_transcripts.txt** - a table of intron to transcript
  relationships. This file is used during the post-processing phase.

## Quantification

**Note**: Technically any quantification tool can be used with `kma`, but
currently only support with `eXpress` is implemented. Please contact me [on
Github](https://github.com/pachterlab/kma/issues) if you're interested in
a quantification tool being supported.

Here, we will discuss quantification. After `generate_introns.py` is run,
`introns.fa` should be combined with the full transcript sequences. This can be
done using the Linux command `cat`:

```{bash}
cat trans.fa introns.fa > trans_and_introns.fa
```

We assume the file name is `trans_and_introns.fa` in the following sections, but
the file name can be anything.

After you've done this, this section requires the following steps:

1. Create the Bowtie 2 index
1. Align reads to the augmented transcriptome
1. Quantify against the augmented transcriptome

### Creating the Bowtie 2 index

See the Bowtie 2 manual for more advanced options.

```{bash}
bowtie2-build --offrate 1 trans_and_introns.fa trans_and_introns
```

This only has to be run once if every time you decide to change the gene
annotation.

### Align reads

Once you have a Bowtie 2 index, you can align any number of RNA-Seq experiments
to that index. The following arguments are recommended when running Bowtie 2:

```{bash}
bowtie2 -k 200 --rdg 6,5 --rfg 6,5 --score-min L,-.6,-.4 -X trans_and_introns
    -1 left.fastq -2 right.fastq | samtools view -Sb - > hits.bam
```

If you don't have many reads, you  might consider replacing the `-k 200`
arugment with `-a`.

### Quantify

Once you've aligned the reads, you can run eXpress against the alignments. See
the eXpress website for additional arguments. The general eXpress call is as
follows:

```{bash}
express trans_and_introns.fa hits.bam
```

If you don't have many reads, consider running additional batch iterations
using the `-B` argument.

## Computing intron retention (post-processing)

The first step is to the load the quantification data into `R`. The function
`read_express` takes a list of file names, sample names, and condition names.
`read_express` then returns a list with the attributes:

- `tpm` - a `data.frame` with TPM of all samples
- `uniq_counts` - a `data.frame` with the number of unique counts of all
  samples
- `all_data` - a list of `data.frame`s from all the eXpress output. Sorted by
  `target_id`.
- `sample` - a character vector for each sample, describing the sample (e.g.
  `tumor1`, `tumor2`)
- `condition` - a character vector for each sample describing the grouping
  (e.g. `tumor`)

This data can then be used by `newIntronRetention`. An example of this can be
seen in [the worked example](#post-processing).

# A worked example

We will run through a small example. The data set including reads [can be found
here](http://math.mcb.berkeley.edu/~pimentel/kma/kma_example.tar.gz). If you
are not interested in the pre-processing and quantification steps (some might
find it simple), the results from the pipeline exist in the R package:

```{r eval=FALSE}
system.file("example", package="kma")
```

In this case, you can simply jump to the [post-processing](#post-processing)
phase which runs through using the R package itself.

This example contains the reads, annotation, and sequences for the SF3B1 gene
in orthochromatic erythroblasts from
[GSE53635](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE53635). This
dataset is unrealistically small, so the results aren't entirely reliable, but
can be run very quickly through the pipeline. Thus, the results aren't very
reliable, but serves as an example of how to run through the pipeline.

## Pre-processing

Run `system.file` in `R` to get the pre-processing directory (see the
pre-processing section above for more information):

```{r, eval=FALSE}
system.file("pre-process", package = "kma")

```

Then, store this value in an environment variable in your shell:

```
PRE=/path/to/kma/pre-process
```

Extract the data from the downloaded archive and move to the example directory and run the pre-processing:

```
tar -xf kma_example.tar.gz
cd kma_example/example
python $PRE/generate_introns.py --genome genome/chr2.fa \
    --gtf annotation/refGene_sf3b1.gtf --extend 25 \
    --out kma_pre-process_out
```

We chose `--extend 25` here because the reads are 51 bases long. An overlap of
25 bases on each side will give us a fair amount of junction reads without
resulting in spurious alignments. If your read length is longer, feel free to
set this value higher.

The output files will then be stored in `kma_pre-process_out` which you can
view using `ls kma_pre-process_out`:

```
intron_to_transcripts.txt introns.bed               introns.fa
```

The `introns.fa` file can now be merged with the transcriptome sequences in the
quantification section.

## Quantification

### Merging the transcriptome and intron sequences

Next, we need to merge the transcriptome and intron sequences. This can be done
using the command `cat`:

```
cat annotation/sf3b1.fa kma_pre-process_out/introns.fa > annotation/sf3b1_with_introns.fa
```

The file `annotations/sf3b1_with_introns.fa` now contains the transcriptome
sequences and the intron sequences. This FASTA file can now be used to build
a Bowtie 2 index for alignment.

### Building the Bowtie 2 index

We want to align against the introns and transcripts. Thus, we will build
a Bowtie 2 index. See the Bowtie 2 website for more information on the
parameters:

```{bash}
bowtie2-build --offrate 1 annotation/sf3b1_with_introns.fa annotation/sf3b1_with_introns
```

This call will result in many Bowtie 2 index files (`*.bt2` files).

### Alignment

For alignment, we will only align one sample as aligning other samples can be
done similarly and can be automated. Please note, the reads we are aligning are
single-end. The command for paired-end reads can be found on the Bowtie
2 website.

```{bash}
bowtie2 -k 200 --rdg 6,5 --rfg 6,5 --score-min L,-.6,-.4 -x annotation/sf3b1_with_introns \
    -U experiment/ortho/ortho1/*.fastq.gz |
    samtools view -Sb - > experiment/ortho/ortho1/sf3b1.bam
```

This will result in a BAM file `experiment/ortho/ortho1/sf3b1.bam` which can
then be quantified using eXpress.

### Quantification with eXpress

Now that we have alignment files, we can quantify isoform and intron expression
with eXpress:

```{bash}
express -o experiment/ortho/ortho1/xprs_out annotation/sf3b1_with_introns.fa \
    experiment/ortho/ortho1/sf3b1.bam
```

The quantification results will now be in `experiment/ortho/ortho1/xprs_out`.
The other samples can be processed similarly.

## Post-processing

### Loading the files

Since we organized our data nicely, we can load the data into R quite easily
using `Sys.glob`:

```{r}
base_dir <- system.file("example", package="kma")
xprs_fnames <- Sys.glob(file.path(base_dir, "experiment/*/*/xprs_out/results.xprs"))
xprs_fnames
```

Sample names can be inferred from the file path as well:

```{r}
sample_names <- sub(file.path(base_dir, "experiment/[a-z]+/"), "", xprs_fnames) %>%
    sub("xprs_out/results.xprs", "", .) %>%
    gsub("/", "", .)
sample_names
```

Since we labelled replicates according to the condition they are a part of, you
can also infer condition names from the sample:

```{r}
condition_names <- sub("[0-9]+", "", sample_names)
condition_names
```

You now have three R variables:

- `xprs_fnames` - a character vector of file names pointing to the
  quantification files
- `sample_names` - a character vector of identifiers for each sample
- `condition_names` - a character vector of identifiers of conditions

We can now load all the eXpress data into R:

```{r}
xprs <- read_express(xprs_fnames, sample_names, condition_names)
```

This results in a named list with the following members:

```{r}
names(xprs)
```

We only need one other file, `intron_to_transcripts.txt`, which was created
during the pre-processing step. We can read that in using `read.table` or
`data.table::fread` which is much faster, just be sure to use the `data.table
= FALSE` flag as seen below:

```{r}
intron_to_trans <- data.table::fread(file.path(base_dir, "kma_pre-process_out",
    "intron_to_transcripts.txt"), data.table = FALSE)
head(intron_to_trans)
```

To create an `IntronRetention` object, you can call `newIntronRetention` with
the following command:

```{r}
ir <- newIntronRetention(xprs$tpm, intron_to_trans, xprs$condition,
    xprs$uniq_counts)
```

`newIntronRetention` take the following arguments:

- expression data frame (in this case we are using TPM from eXpress)
- intron to transcripts relationship table
- condition character vector
- Optional: a unique counts vector. This is used for filtering later on

See `help(newIntronRetention)` for more details on the arguments.

Printing the `IntronRetention` object results in a short summary:

```{r}
print(ir)
```

The data member `flat` in the `IntronRetention` object is where most of the
operations happen. Advanced users might be interested in performing direct
exploratory analysis on this table, which is similar to a denormalized SQL
table (often reffered to as a "tidy" table in R).

```{r}
head(ir$flat)
```

### Filtering

Filter functions have the following format: `filter_name(ir, options)` where
`ir` is an `IntronRetention` object, and options are options for the filter.
They return an `IntronRetention` object with an updated `flat` member. Filters
can be implemented by the user, but should return `TRUE` if the intron passes
the filter and `FALSE` otherwise. The column name for each filter should begin
with `f_`.

These filters below filter denominators that have expression below 1 TPM,
introns that have exactly 0 or 1 PSI (due to either no contributions from the
intron coverage or solely coverage from the intron). `filter_low_frags` filters
out introns that don't have at least `N` number of unique reads:

```{r}
ir <- ir %>%
    filter_low_tpm(1) %>%
    filter_perfect_psi() %>%
    filter_low_frags(3)
colnames(ir$flat)
```

#### Zero coverage

The zero coverage filter is a special type of filter that has to be computed on
the read alignments and corresponding eXpress results.

```{bash}
python $PRE/zeroCoverage.py experiment/ortho/ortho1/xprs_out/results.xprs \
    experiment/ortho/ortho1/sf3b1.bam \
    experiment/ortho/ortho1/zero_coverage.txt
```

The output is then in `experiment/ortho/ortho1/zero_coverage.txt`. Zero
coverage data can be read in using the function `get_batch_intron_zc`. Like
`read_express` it needs file names, sample names, and condition names:

```{r}
zc_fnames <- Sys.glob(file.path(base_dir, "experiment/*/*/zero_coverage.txt"))
zc_samples <- sub(file.path(base_dir, "experiment/[a-z]+/"), "", zc_fnames) %>%
    sub("zero_coverage.txt", "", .) %>%
    gsub("/", "", .)
zc_conditions <- sub("[0-9]+", "", zc_samples)
all_zc <- get_batch_intron_zc(zc_fnames, zc_samples, zc_conditions)
head(all_zc)
```

The result from `get_batch_intron_zc` is a `data.frame` that can be inspected
manually if you'd like. To incorporate it with a `IntronRetention` object, use
`summarize_zero_coverage`:

```{r}
ir <- summarize_zero_coverage(ir, all_zc)
```

This adds a new column to `ir$flat` called `f_zc_*`:

```{r}
colnames(ir$flat)
```

### Hypothesis testing

Hypothesis testing aggregates all filters (by taking the intersection of all
filters), computes the null distribution, then returns a data frame summarizing
the test. Since there is some randomness involved in generating the
permutations, you might consider setting a seed before calling `retention_test`
as we do below:

```{r}
set.seed(42)
ir_test <- retention_test(ir)
head(ir_test)
```

Significant introns can be found using the dplyr function `filter`:

```{r}
ir_test %>%
    filter(qvalue <= 0.10) %>%
    select(-c(pvalue))
```

In this case we only see one significant intron. Since we have such a small
dataset, the null distribution isn't that reliable so the results shouldn't be
taken too seriously, but when the data set is larger (like in a normal
experiment), the results are more reliable.


# Intron retention in terminal erythropoiesis

**TODO** This section is coming soon. It will showcase how to make plots as
well as some diagnostics one might do on a complete data set.
